Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equations

### 2014-03-27 pituba/article/article.tex
Changes in text

Oil component is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature (stock tank oil). In contrast, gas component is composed by hydrocarbon species in gaseous phase for the same surface conditions.

Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component. There can exist three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by the gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

@@

We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, also known as formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature conditions to the volume of this same phase in standard pressure and temperature condition, i.e.:

@@

Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations. So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.

@@

\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1}
\end{equation}
If fluid densities do not vary appreciately with depth.

@@

A single equation for pressure (without fluid saturations) can be obtained by substituting Eqs.~\eqref{eq: So2}, \eqref{eq: Sg2} and \eqref{eq: Sw2} into \eqref{eq: S12}. The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}

>>>>

Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature. Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions. All species are generally present in both phases although the equilibrium composition is shifted to liquid phase (or gaseous phase) for components of greater (lower) molecular mass.

Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component. There can exist three fluid phases: oil, gas and liquid water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by the gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

@@

We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, also known as formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature condition to the volume of this same phase in standard pressure and temperature, i.e.:

@@

Performing time integration, balance equations for water and oil are:
\begin{equation}\label{eq: So5}
\frac{\bar{S}_{w1} V_{p1}}{B_{w1}} - \frac{\bar{S}_{w2} V_{p2}}{B_{w2}} = W_p \, ,
\end{equation}
\begin{equation}\label{eq: So5}
\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = \left(N-N_p\right) \, ,
\end{equation}
where $N=\bar{S_{o1}} V_{p1} / B_{o1}$ is the initial volume of oil measured in standard conditions.

Balance of gas component is:
\begin{equation}\label{eq: Sg5}
V_{p2}\left(\frac{S_{g2}}{B_{g2}} +\frac{S_{o2} R_{s2}}{B_{o2}}\right) - N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right)=G_i-G_p \, ,
\end{equation}
where $m=V_{p1} S_{g1}/ B_{o1} N$ is the ratio of initial volumes of gas and oil phases in reservoir conditions.

%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations. So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.

@@

\begin{equation}
\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1}
\end{equation}
If fluid densities do not vary appreciably with depth.

@@

A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So2}, \eqref{eq: Sg2} and \eqref{eq: Sw2} into \eqref{eq: S12}. The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}
